---
title: "Data 605 HW8: Sums of RVs, Law of Large Numbers"
author: "Sin Ying Wong"
date: "10/18/2020"
output:
  rmdformats::readthedown:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    smooth_scroll: yes
    theme: united
    toc_collapsed: yes
    toc_depth: 5
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
    toc_collapsed: yes
    toc_float: yes
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
  word_document:
    toc: yes
    toc_depth: '5'
theme: lumen
number_sections: yes
toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

Assignment 8: Introduction to Probability Textbook page303 Sec7.2 Ex11, Ex14, and page320 Sec8.2 Ex1.

```{r libraries, message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(gtools)
```

# Problem Set 1 (Sec7.2 Ex11)

A company buys 100 lightbulbs, each of which has an exponential lifetime of 1000 hours. 
What is the expected time for the first of these bulbs to burn out? (See Exercise 10.)

## PS1 Answer

From Ex10, we know that, let $X_{1}, X_{2}, \cdots, X_{n}$ be $n$ independent random variables each of which has an exponential density with mean $\mu$ and let $M$ be the minimum value of the $X_{j}$, the density for $M$ is exponential with mean $\frac{\mu}{n}$.

By using this property, when $X_{i}\sim Exp(\lambda_{i})$, we have $min(X_{i})\sim Exp(\Sigma \lambda_{i})$

For Ex11, from the given information, we have $n=100$ exponential variables where each represents the lifetime of one lightbulb with the average as $1000$ hours.  Assuming independence, the expected time for the first lightbulb to burnout is the minimum of lifetime of this set of $100$ lightbulbs.

For any individual lightbulb, mean $=1000$ and $E[x]=\frac{1}{\lambda}=\frac{1}{1000}$

For $min(X_{i})$, new mean $= mean^{'}=\frac{mean}{n}=\frac{1000}{100}=10=\frac{1}{\lambda^{'}}$. Thus, $\lambda^{'}=\frac{1}{10}$ and $E[min(X_{i})]=\frac{1}{\lambda^{'}}=10$.

Therefore, the expected time for the first of these 100 lightbulbs to burn out is 10 hours.


# Problem Set 2 (Sec7.2 Ex14)

Assume that $X_{1}$ and $X_{2}$ are independent random variables, each havng an exponential density with parameter $\lambda$. 

Show that $Z = X_{1}-X_{2}$ has density $f_{Z}(z)=\frac{1}{2}\lambda e^{-\lambda \left | z \right |}$.

## PS2 Answer

As both $X_{1}$ and $X_{2}$ are independent exponential variables, both are evaluated on the interval $[0,\infty)$.

We have $f_{X_{1}}(x_{1})=f_{X_{2}}(x_{2})= \left\{\begin{matrix}\lambda e^{-\lambda x} & x \geq 0\\ 0 & otherwise \end{matrix}\right.$

$$f_{Z}(z)= f_{X_{1}+(-X_{2})}(z) = \int_{-\infty}^{\infty} f_{x_{1}}(x_{1})f_{-x_{2}}(z-x_{1})dx_{1}$$
$$=\int_{-\infty}^{\infty} f_{x_{1}}(x_{1})f_{x_{2}}(x_{1}-z)dx_{1}$$
$$=\int_{-\infty}^{\infty} \lambda e^{-\lambda (x_{1})} \lambda e^{-\lambda (x_{1}-z)} dx_{1}$$
$$=\int_{-\infty}^{\infty} \lambda^{2} e^{\lambda (z-2x_{1})} dx_{1}$$

For $z \geq 0$ and $x_{2}=x_{1}-z \geq 0,$ then $x_{1}\geq z\,$. For $z<0$ and $x_{2}=x_{1}-z \geq 0,$ then $x_{1}\geq 0$.


- For $z<0,\, x_{1}\geq 0$,

$$f_{Z}(z)= f_{X_{1}+(-X_{2})}(z) = \int_{-\infty}^{\infty} f_{x_{1}}(x_{1})f_{x_{2}}(x_{1}-z)dx_{1}$$
$$=\int_{-\infty}^{\infty} \lambda^{2} e^{\lambda (z-2x_{1})} dx_{1}$$
$$= \int_{0}^{\infty} \lambda e^{\lambda z} \lambda e^{-2\lambda x_{1}} dx_{1}$$
$$=-\frac{\lambda}{2} e^{\lambda z}  \int_{0}^{\infty} -2\lambda e^{-2\lambda x_{1}} dx_{1}$$
$$=-\frac{\lambda}{2} e^{\lambda z} \left [ e^{-2\lambda x_{1}} \right ]^{\infty}_{0}$$
$$=-\frac{\lambda}{2} e^{\lambda z}[0-1]$$
$$=\frac{\lambda}{2} e^{\lambda z}$$


- For $z \geq 0,\, x_{1}\geq z$,

$$f_{Z}(z)= f_{X_{1}+(-X_{2})}(z) = \int_{-\infty}^{\infty} f_{x_{1}}(x_{1})f_{x_{2}}(x_{1}-z)dx_{1}$$
$$=\int_{-\infty}^{\infty} \lambda^{2} e^{\lambda (z-2x_{1})} dx_{1}$$

$$= \int_{z}^{\infty} \lambda e^{\lambda z} \lambda e^{-2\lambda x_{1}} dx_{1}$$
$$=-\frac{\lambda}{2} e^{\lambda z}  \int_{z}^{\infty} -2\lambda e^{-2\lambda x_{1}} dx_{1}$$
$$=-\frac{\lambda}{2} e^{\lambda z} \left [ e^{-2\lambda x_{1}} \right ]^{\infty}_{z}$$
$$=-\frac{\lambda}{2} e^{\lambda z}[0-e^{-2\lambda z}]$$
$$=\frac{\lambda}{2} e^{-\lambda z}$$

Combine the above results, we have

$$f_{Z}(z)= \left\{\begin{matrix} \frac{\lambda}{2} e^{\lambda z} & z < 0\\ \frac{\lambda}{2} e^{-\lambda z} & z\geq 0 \end{matrix}\right.$$
$$f_{Z}(z)=\frac{1}{2}\lambda e^{-\lambda \left | z \right |}$$



# Problem Set 3 (Sec8.2 Ex1)

Let $X$ be a continuous random variable with mean $\mu =10$ and variance $\sigma ^{2} = \frac{100}{3}$. Using Chebyshev's Inequality, find an upper bound for the following probabilities.

(a.) $P(\left | X-10 \right | \geq 2)$

(b.) $P(\left | X-10 \right | \geq 5)$

(c.) $P(\left | X-10 \right | \geq 9)$

(d.) $P(\left | X-10 \right | \geq 20)$

## PS3 Answer

From textbook page306 Example 8.1, we have the following information:

Let $X$ by any random variable with $E(X)=\mu$ and $V(X)=\sigma^{2}$. Then, if $\epsilon=k\sigma$, Chebyshev's Inequality states that
$$P(\left | X-\mu \right | \geq k\sigma) \leq \frac{\sigma^{2}}{k^{2}\sigma^{2}}=\frac{1}{k^{2}}$$

### Part a

By using Chebyshev’s Inequality and the given information $\mu=10$ and $\sigma^{2}=\frac{100}{3}$, we have $\sigma=\frac{10}{\sqrt3}$.

For $k\sigma=2$, we have $k=2\times \frac{\sqrt3}{10}$.

$$\because \frac{1}{k^{2}}=\frac{100}{2^{2}\times3}=\frac{25}{3}>1 \;\; and \;\; max(Probability)=1$$

$$\therefore P(\left | X-10 \right | \geq 2) < 1$$


### Part b

By using Chebyshev’s Inequality and the given information $\mu=10$ and $\sigma^{2}=\frac{100}{3}$, we have $\sigma=\frac{10}{\sqrt3}$

For $k\sigma=5$, we have $k=5\times \frac{\sqrt3}{10}$.

$$\because \frac{1}{k^{2}}=\frac{100}{5^{2}\times3}=\frac{4}{3}>1 \;\; and \;\; max(Probability)=1$$

$$\therefore P(\left | X-10 \right | \geq 5) < 1$$

### Part c

By using Chebyshev’s Inequality and the given information $\mu=10$ and $\sigma^{2}=\frac{100}{3}$, we have $\sigma=\frac{10}{\sqrt3}$

For $k\sigma=9$, we have $k=9\times \frac{\sqrt3}{10}$.

$$\therefore P(\left | X-10 \right | \geq 9) \leq \frac{1}{k^{2}}=\frac{100}{81\times3}=\frac{100}{243}$$

### Part d

By using Chebyshev’s Inequality and the given information $\mu=10$ and $\sigma^{2}=\frac{100}{3}$, we have $\sigma=\frac{10}{\sqrt3}$

For $k\sigma=20$, we have $k=20\times \frac{\sqrt3}{10}$.

$$\therefore P(\left | X-10 \right | \geq 20) \leq \frac{1}{k^{2}}=\frac{100}{400\times3}=\frac{1}{12}$$