---
title: "Data 605 HW2: Trace, Determinant, Factorization"
author: "Sin Ying Wong"
date: "09/06/2020"
output:
  rmdformats::readthedown:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    smooth_scroll: yes
    theme: united
    toc_collapsed: yes
    toc_depth: 5
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
    toc_collapsed: yes
    toc_float: yes
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
  word_document:
    toc: yes
    toc_depth: '5'
theme: lumen
number_sections: yes
toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

Please refer to the [Assignment 2 Document](https://github.com/shirley-wong/Data-605/blob/master/Assignment2/hw2.pdf).

```{r libraries, message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(pracma)
```

# Problem Set 1

(1) Show that $A^{T}A \neq AA^{T}$ in general. (Proof and demonstration.)
(2) For a special type of square matrix A, we get $A^{T}A = AA^{T}$ . Under what conditions
could this be true? (Hint: The Identity matrix I is an example of such a matrix).

Please typeset your response using LaTeX mode in RStudio. If you do it in paper, please
either scan or take a picture of the work and submit it. Please ensure that your image is
legible and that your submissions are named using your first initial, last name, assignment
and problem set within the assignment. E.g. LFulton_Assignment2_PS1.png

## Part(1)
Q: Show that $A^{T}A \neq AA^{T}$ in general. (Proof and demonstration.)

### Proof
A: Suppose a matrix $A$ is an $m \times n$ matrix and a matrix $B$ is an $n \times p$ matrix. Let vectors $A_{1}, A_{2}, A_{3},...,A_{n}$ denote the columns of $A$ and let the vectors $B_{1}, B_{2}, B_{3},...,B_{p}$ denote the columns of $B$. 

Then for $1\leq i\leq m, 1\leq j\leq p$, the individual entries of $AB$ are:


$$[AB]_{ij} = [AB_{j}]_{i}$$
$$= [ [B_{j}]_{1}A_{1} + [B_{j}]_{2}A_{2} + \cdots + [B_{j}]_{n}A_{n} ]_{i}$$
$$= [ [B_{j}]_{1}A_{1}]_{i} + [[B_{j}]_{2}A_{2}]_{i} + \cdots + [[B_{j}]_{n}A_{n} ]_{i}$$
$$= [B_{j}]_{1}[A_{1}]_{i} + [B_{j}]_{2}[A_{2}]_{i} + \cdots + [B_{j}]_{n}[A_{n}]_{i}$$
$$= [B_{1j}][A]_{i1} + [B_{2j}][A]_{i2} + \cdots + [B_{nj}][A]_{in} $$
$$= [A_{i1}][B]_{1j} + [A_{i2}][B]_{2j} + \cdots + [A_{in}][B]_{nj}$$
$$= \sum_{k=1}^{n} [A]_{ik}[B]_{kj}$$

Base on the proof above, suppose the matrix $A$ is an $m \times  n$ matrix, which implies that $A^{T}$ is an $n \times m$ matrix. Then for $1\leq i\leq m, 1\leq j\leq m$, the individual entries of $AA^{T}$ are given by 
$$[AA^{T}]_{ij} = [A]_{i1}[A^{T}]_{1j} + [A]_{i2}[A^{T}]_{2j} + [A]_{i3}[A^{T}]_{3j} + \cdots + [A]_{in}[A^{T}]_{nj}$$
$$= \sum_{k=1}^{n} [A]_{ik}[A^{T}]_{kj}$$


And for $1\leq i\leq n, 1\leq j\leq n$, the individual entries of $A^{T}A$ are given by 
$$[A^{T}A]_{ij} = [A^{T}]_{i1}[A]_{1j} + [A^{T}]_{i2}[A]_{2j} + [A^{T}]_{i3}[A]_{3j} + \cdots + [A^{T}]_{im}[A]_{mj}$$
$$= \sum_{k=1}^{m} [A^{T}]_{ik}[A]_{kj}$$
, where $AA^{T}$ is an $m \times m$ matrix and $A^{T}A$ is an $n \times n$ matrix.

Therefore, $A^{T}A \neq AA^{T}$ in general.


### Demonstration
Let $A = \begin{bmatrix} 1 & 4\\ 2 & 5\\ 3 & 6\end{bmatrix}$ and $A^{T} = \begin{bmatrix} 1 & 3 & 5\\ 2 & 4 & 6\end{bmatrix}$, and calculate $A^{T}A$ and $AA^{T}$.


```{r Q1p1}
A <- matrix(c(1,2,3,4,5,6), nrow=3, ncol=2)
A
A_t <- t(A)
A_t

A_t%*%A

A%*%A_t
```

By the calculation above, we have $A^{T}A = \begin{bmatrix} 14 & 32\\ 32 & 77\end{bmatrix}$ and $AA^{T} = \begin{bmatrix} 17 & 22 & 27\\ 22 & 29 & 36\\ 27 & 36 & 45\end{bmatrix}$, which shows that $A^{T}A \neq AA^{T}$ in general.



## Part(2)
Q: For a special type of square matrix A, we get $A^{T}A = AA^{T}$ . Under what conditions
could this be true? (Hint: The Identity matrix I is an example of such a matrix).

A: Based on the proof in Part(1), we have two equations:
$$[A^{T}A]_{ij} = \sum_{k=1}^{m} [A^{T}]_{ik}[A]_{kj}$$
$$[AA^{T}]_{ij} = \sum_{k=1}^{n} [A]_{ik}[A^{T}]_{kj}$$

In order to have $A^{T}A = AA^{T}$, we need $m=n$ and $A = A^{T}$, which implies $A$ has to be a symmetric matrix.


# Problem Set 2

Matrix factorization is a very important problem. There are supercomputers built just
to do matrix factorizations. Every second you are on an airplane, matrices are being
factorized. Radars that track flights use a technique called Kalman filtering. At the heart
of Kalman Filtering is a Matrix Factorization operation. Kalman Filters are solving linear
systems of equations when they track your flight using radars.

Write an R function to factorize a square matrix A into LU or LDU, whichever you
prefer. Please submit your response in an R Markdown document using our class naming
convention, E.g. LFulton_Assignment2_PS2.png

You don't have to worry about permuting rows of A and you can assume that A is less
than 5x5, if you need to hard-code any variables in your code. If you doing the entire
assignment in R, then please submit only one markdown document for both the problems.

## LDU Decomposition

To create a function for LDU decomposition, we have a few steps to do.

1. Find the dimension of the given matrix $A$ ($m \times n$ matrix) and make sure it is a square matrix, i.e. m=n.
2. Initialize the three square matrices $L$, $D$, and $U$.
3. Do row operations on $A$ to find $L$. The matrix $A$ here will become an upper-triangular matrix while $L$ is a lower-triangular matrix.
4. Continue to do row operations on $A$ to find $D$, a diagonal matrix, and $U$, an upper triangular matrix.
5. Verify my function LDU by multiplying the matrices $L$, $D$, and $U$ and see if it equals to the original matrix $A$.

### LDU Decomposition Function
```{r LDU}
LDU <- function(a){
  A <- a
  if (nrow(A) != ncol(A)){
    p <- ("Your matrix is not eligible for LDU Decomposition. Please input a square matrix.")
    return(p)
  }
  else{
    #initialize the three matrices L, D, and U
    L <- diag(nrow(A))
    D <- diag(nrow(A))
    U <- diag(nrow(A))
    
    #row operations on A to reduce A into an upper-triangular matrix and find L
    for (n in 1:ncol(A)){
      for (m in 2:nrow(A)){
        if (m > n){
          L[m,n] <- A[m,n]/A[n,n]
          A[m,] <- A[m,] - A[m,n]/A[n,n]*A[n,]
        }
      }
    }
    
    #row operations on the upper-triangular matrix A to find D and U
    for (n in 1:ncol(A)){
      for (m in 1:nrow(A)){
        if (m == n){
          D[m,n] = A[m,n]
          A[m,] = A[m,]/A[m,n]
        }
      }
    }
    
    #the A after all operations is indeed U
    U <- A
    
    return(list('A'=a, 'L'=round(L,3), 'D'=round(D,3), 'U'=round(U,3), 'LDU'=round((L%*%D%*%U),3), 'A equals LDU' = all((round(a,3)==round(L%*%D%*%U,3)))))
  }
}


```

### Test the function

1. Test with a 3x2 matrix $A = \begin{bmatrix} 3 & 10\\ 8 & -3\\ -1 & 2\end{bmatrix}$
```{r test_LDU_3x2}
A <- matrix(c(3,10,8,-3,-1,-2), nrow=3, ncol=2, byrow=TRUE)
LDU(A)
```

2. Test with a 3x3 matrix $A = \begin{bmatrix} 3 & 10 & 7\\ 8 & -3 & 5\\ -1 & -2 & 4\end{bmatrix}$
```{r test_LDU_3x3}
A <- matrix(c(3,10,7,8,-3,5,-1,-2,4), nrow=3, ncol=3, byrow=TRUE)
LDU(A)
```


3. Test with a 6x6 matrix $A = \begin{bmatrix}3 & -9 & -13 & 22 & 18 & -24\\ -25 & 21 & -12 & -7 & 23 & -27\\ 12 & -24 & 27 & -8 & 24 & 16\\ 9 & -4 & -9 & 17 & 26 & -22\\ 21 & -29 & 11 & -30 & 10 & -31\\ -21 & 3 & 26 & -14 & 37 & 19\end{bmatrix}$
```{r test_LDU_6x6}
A <- matrix(c(3,-9,-13,22,18,-24,-25,21,-12,-7,23,-27,12,-24,27,-8,24,16,9,-4,-9,17,26,-22,21,-29,11,-30,10,-31,-21,3,26,-14,37,19), nrow=6, ncol=6, byrow=TRUE)
LDU(A)
```