---
title: "Data 605 HW7: Imp. Distributions, EX, VARX"
author: "Sin Ying Wong"
date: "10/10/2020"
output:
  rmdformats::readthedown:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    smooth_scroll: yes
    theme: united
    toc_collapsed: yes
    toc_depth: 5
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
    toc_collapsed: yes
    toc_float: yes
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
  word_document:
    toc: yes
    toc_depth: '5'
theme: lumen
number_sections: yes
toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

Please refer to the [Assignment 7 Document](https://github.com/shirley-wong/Data-605/blob/master/Assignment7/hw7.pdf).

```{r libraries, message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(gtools)
```

# Problem Set 1 (Sec5.1 Ex6)

Let $X_{1}, X_{2}, \cdots , X_{n}$ be $n$ mutually independent random variables, each of which is uniformly distributed on the integers from $1$ to $k$. Let $Y$ denote the minimum of the $X_{i}$'s. Find the distribution of $Y$.


## PS1 Answer

Given that $i \in [1,n], \;\forall X_{i} \in [1,k], \;Y=min(X_{i})$

For $Y=1,$ the probability for an $X_{i}=1$ is $\frac{1}{k}$ and the probability for an $X_{i}>1$ is $\frac{k-1}{k}$.

$$\therefore P(Y=1)=P(min (X_{i})=1)=1-(\frac{k-1}{k})^{n}$$
$$=(\frac{k-0}{k})^{n}-(\frac{k-1}{k})^{n}$$

For $Y=2,$ the probability for an $X_{i}>2$ is $\frac{k-2}{k}$.

$$\therefore P(Y=2)=P(min (X_{i})=2)=1-P(Y=1)-P(Y>2)$$
$$=1-(1-(\frac{k-1}{k})^{n})-(\frac{k-2}{k})^{n}$$
$$=(\frac{k-1}{k})^{n}-(\frac{k-2}{k})^{n}$$

For $Y=3,$ the probability for an $X_{i}>3$ is $\frac{k-3}{k}$.

$$\therefore P(Y=3)=P(min (X_{i})=3)=1-P(Y=1)-P(Y=2)-P(Y>3)$$
$$=1-(1-(\frac{k-1}{k})^{n})-((\frac{k-1}{k})^{n}-(\frac{k-2}{k})^{n})-(\frac{k-3}{k})^{n}$$
$$=(\frac{k-2}{k})^{n}-(\frac{k-3}{k})^{n}$$

Therefore, by induction, the distribution of Y is
$$P(Y=y)=P(min (X_{i})=y)=(\frac{k-y+1}{k})^{n}-(\frac{k-y}{k})^{n}\;\;\;for\,y \in[1,k]\,,\;k\in \mathbb{Z}^{+}$$



# Problem Set 2

Your organization owns a copier (future lawyers, etc.) or MRI (future doctors). This machine has a manufacturer's expected lifetime of 10 years. This means that we expect one failure every ten years. (Include the probability statements and R Code for each part.).

(a.) What is the probability that the machine will fail after 8 years? Provide also the expected value and standard deviation. Model as a geometric. (Hint: the probability is equivalent to not failing during the first 8 years.)

(b.) What is the probability that the machine will fail after 8 years? Provide also the expected value and standard deviation. Model as an exponential.

(c.) What is the probability that the machine will fail after 8 years? Provide also the expected value and standard deviation. Model as a binomial. (Hint: 0 success in 8 years.)

(d.) What is the probability that the machine will fail after 8 years? Provide also the expected value and standard deviation. Model as a Poisson.


## PS2 Answer

```{r Q2 general, message=FALSE, warning=FALSE}
lifetime = 10
p = 1/lifetime #probability of machine failure
q = 1-p
n = 8
```

### a. Geometric

Assuming the unit is annual. Geometric pmf is $p(1-p)^{n-1}$, $E[X]=\frac{1}{p}$ and $V[X]=\frac{1-p}{p^{2}}$.

The probability that the machine will fail after 8 years is $0.4304672$.

The expected value is $10$ and the standard deviation is $9.486833$.

```{r Q2a, message=FALSE, warning=FALSE}
q^8 #probability of 0 machine failures in 8 years
pgeom(n-1,p,lower.tail = FALSE) 

1/p #expected value

sqrt((1-p)/(p^2)) #sd
```

### b. Exponential

Exponential pdf is $\lambda e^{-\lambda x}$ for $x\geq0$ and 0 otherwise. $E[X]=\frac{1}{\lambda}$ and $V[X]=\frac{1}{\lambda^{2}}$, where $\lambda=1/10=p$.

The probability that the machine will fail after 8 years is $0.449329$.

The expected value is $10$ and the standard deviation is $10$.

```{r Q2b, message=FALSE, warning=FALSE}
pexp(n, p, lower.tail = FALSE)

1/p #E[X]

sqrt(1/p^2) #sd
```

### c. Binomial

Binomial pmf is $\begin{pmatrix}n\\ p\end{pmatrix}p^{k}(1-p)^{n-k}$. $E[X]=np$ and $V[X]=np(1-p)$.

The probability that the machine will fail after 8 years is $0.4304672$.

The expected value is $0.8$ and the standard deviation is $0.8485281$.

```{r Q2c, message=FALSE, warning=FALSE}
pbinom(0,n,p) #0 machine failures in 8 years

n*p #E[X]

sqrt(n*p*q) #sd
```

### d. Poisson

Poisson pmf is $\frac{\lambda^{k}e^{-\lambda}}{k!}$ where $k$ is the number of occurrences. $E[X]=\lambda$ and $V[X]=\lambda$.

As the average number of machine failure in every 10 years is 1, the average number of occurrences in 8 years will be 8/10.

The probability that the machine will fail after 8 years is $0.449329$.

The expected value is $0.8$ and the standard deviation is $0.8944272$.

```{r Q2d, message=FALSE, warning=FALSE}
lambda = n/lifetime #the average number of failures in n years

ppois(0, lambda) #0 machine failures in 8 years

lambda #E[X]

sqrt(lambda) #sd
```

