---
title: "Data 605 HW11: Regression Analysis in R 1"
author: "Sin Ying Wong"
date: "11/08/2020"
output:
  rmdformats::readthedown:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    smooth_scroll: yes
    theme: united
    toc_collapsed: yes
    toc_depth: 5
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
    toc_collapsed: yes
    toc_float: yes
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
  word_document:
    toc: yes
    toc_depth: '5'
theme: lumen
number_sections: yes
toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

# Regression Problem

Using the “cars” dataset in R, build a linear model for stopping distance as a function of speed and replicate the analysis of your textbook chapter 3 (visualization, quality evaluation of the model, and residual analysis.)

## Import Data

There are 50 observations of speed of cars and the distances taken to stop from the dataset `cars` without any NA values.

```{r import data}
data <- datasets::cars
head(cars)
summary(cars)
dim(data)
```

## Visualize the Data

```{r scatter}
plot(data, main = "Stopping Dist vs Speed", xlab = "Speed (mph)", ylab = "Stopping Distance (ft)")
```

## Build a Linear Model

Build a simple linear model with the dataset `cars`.

```{r lm}
cars_lm <- lm(dist~speed, data=cars)
plot(data, main = "Stopping Dist vs Speed", xlab = "Speed (mph)", ylab = "Stopping Distance (ft)")
abline(reg=cars_lm, col="blue")
```

## Evaluate the Model

Quality evaluation of the model:

Residuals: the median is close to zero and the 1Q vs 3Q and min vs max are roughly the same magnitude. This shows that the model is a fairly good model.

P-value: The p-value of speed is very close to zero, which shows a strong significance at 99% confidence, while the intercept is significant at 95% confidence. This shows that the model is predicting the two values quite well.

Multiple R-squared: The $R^{2}$ value represents how well the model describes the measured data. It explains 65.11% of the data's variation, which is fairly good.

```{r evaluate}
summary(cars_lm)
```

## Residual Analysis

The residuals vary around and are not uniformly scattered above and below zero. They do not show any special patterns.

```{r}
plot(fitted(cars_lm), resid(cars_lm), 
     main = "Cars: Fitted Stopping Distance vs Residuals", 
     xlab = "Fitted Stopping Distance", ylab = "Residuals")
abline(h=0, col="blue")
```

The QQ plot shows some outliers above the line at the upper end and the lower end. The residuals might not be very normal.

```{r}
qqnorm(resid(cars_lm))
qqline(resid(cars_lm))
```

To check the normality, we can also look at the histogram or a normal probability plot of the residuals.

The histogram is bimodal and slightly right-skewed. Therefore, the residuals are barely normal.
```{r}
hist(resid(cars_lm), breaks=15, prob=TRUE)
curve(dnorm(x, mean=mean(resid(cars_lm)), sd=sd(resid(cars_lm))), col="red", add=TRUE)
```
